---
title: "Quick Intro to R"
output:
  pdf_document:
    toc: True
  html_document:
    toc: True
    df_print: paged
---

# Welcome to R
In this very brief introduction to R, we work up to a few examples from two popular packages: *dplyr* for data wrangling and *ggplot2* for plotting. Additionally, we give two examples of some common statistical models. We will use this code later in an interactive Shiny application. These topics comprise some of R's greatest strengths:
+ Data wrangling
+ Plotting/visuals
+ Statistical model fitting
+ Interactive applications (Shiny)
+ All in a fantastic GUI (RStudio)

**You can pull this entire repo down from GitHub using this url: https://github.com/milliman/ADA_IntroToR_Shiny.git **

# Packages
```{r, message = F, warning = F}
# install.packages(c("dplyr", "ggplot2"))
library(dplyr)
library(ggplot2)
```
```{r}
sessionInfo()$R.version$version.string
sapply(sessionInfo()$otherPkgs, function(x) x$Version)
```

# Objects

## Scalars
```{r}
1 + 1
a <- 1 + 1
```

## Vectors
We show various ways to make similar numeric vectors:
```{r}
v1 <- 1:3
v2 <- c(1, 2, 3)
v3 <- seq(1, 3, by = 1)
v4 <- seq(1, 3, length.out = 3)
v1
v2
v3
v4
v1*v2
```

## Character and factor vectors
```{r}
c1 <- c("one", "two", "three", "one", "two", "three")
c1
f1 <- factor(c("one", "two", "three", "one", "two", "three"))
f1
```

## Matrices
```{r cars}
m1 <- matrix(data = c(1, 2, 3, 4, 5, 6, 7, 8, 9), 
             nrow = 3, 
             ncol = 3,
             byrow = FALSE)

m2 <- matrix(data = rnorm(n = 9, mean = 0, sd = 1),
             nrow = 3,
             ncol = 3,
             byrow = TRUE)
m1 * m2
m1 %*% m2
```

## Data frame objects
```{r}
d <- data.frame(A = 1:4,
                B = c("red", "blue", "yellow", "green"))
class(d)
class(iris)
summary(iris)
str(iris)
head(iris)
```

# Reading and writing data
## CSV
```{r}
write.csv(x = iris, 
          file = "iris.csv",
          row.names = F)
mattysdf <- read.csv(file = "iris.csv")
```

## RDS
```{r}
saveRDS(object = iris,
        file = "iris.RDS")
mattysdf <- readRDS("iris.RDS")
```

# Data wrangling
We turn to our favorite data wrangling package, dplyr. 

## Select specific columns
```{r}
mattysdf %>%
  select(Petal.Length, Species) %>%
  head(5)
```

## Filter to specific rows
```{r}
mattysnewdf <- mattysdf %>%
  filter(Species == "versicolor")
```

## "Mutate" on new columns
```{r}
mattysnewdf2 <- mattysdf %>%
  mutate(Sepal.Area = Sepal.Width*Sepal.Length,
         Petal.Area = Petal.Width*Petal.Length)
```

## Pivot table
```{r}
mattysnewdf2 %>%
  group_by(Species) %>%
  summarize(Sepal.Area.mean = mean(Sepal.Area),
            Sepal.Area.sd = sd(Sepal.Area),
            Petal.Area.mean = mean(Petal.Area),
            Petal.Area.sd = sd(Petal.Area))
```

# Plotting
## Base R
### Scatter plot
```{r}
plot(x = mattysnewdf2$Petal.Area,
     y = mattysnewdf2$Sepal.Area)

plot(x = mattysnewdf2$Petal.Area,
     y = mattysnewdf2$Sepal.Area,
     col = ifelse(mattysnewdf2$Species == "setosa", "red",
                  ifelse(mattysnewdf2$Species == "versicolor", "green", "blue")))
```


### Histogram
```{r}
hist(x = faithful$waiting,
     breaks = 20)

hist(x = faithful$waiting,
     breaks = seq(min(faithful$waiting), max(faithful$waiting), length.out = 20 + 1))
```

## ggplot
### Scatter plot
```{r}
mattysnewdf2 %>%
  ggplot(mapping = aes(x = Petal.Area, y = Sepal.Area)) +
  geom_point(mapping = aes(color = Species))

# More cooler scatter plot
mattysnewdf2 %>%
  ggplot(mapping = aes(x = Petal.Area, y = Sepal.Area)) +
  geom_point(mapping = aes(color = Species)) +
  geom_smooth(method = "lm")
```

### Histogram
```{r}
faithful %>%
  ggplot(mapping = aes(x = waiting)) +
  geom_histogram(bins = 20)

# More betterer histogram
mattysnewdf2 %>%
  ggplot(mapping = aes(x = Petal.Area)) +
  geom_histogram(mapping = aes(fill = Species),
                 bins = 20)
```

# Statistical models
R has a plethora of functions to create statistical models. Below we share two of the classics.

## OLS
```{r}
faithful.model <- lm(formula = eruptions ~ waiting,
                     data = faithful)
summary(faithful.model)
```

## Logistic GLM
```{r}
iris.model <- glm(formula = versicolor ~ Sepal.Area + Petal.Area,
                  data = mattysnewdf2 %>%
                    mutate(versicolor = ifelse(Species == "versicolor", 1, 0)),
                  family = binomial(link = "logit"))
summary(iris.model)
```

